An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-03-29 EconGrowthSeminar.Rnw
Now including tables and frames for sources of economic data. Also including graphics for world bank and BLS.

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item{$\sim$10\%}
    \vspace{\baselineskip}
    \pause
    \item{Energy (relatively) unimportant for classical economists}
    \item{Usually ignored}
    \vspace{\baselineskip}
    \pause
    \item{But \dots}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Growth_Models.pdf}
  \end{center}
  \begin{itemize}
    \item{Goal: predict economic output over time}
    \item{Economic output ($Y$) is a function of ``factors of production''}
    \item{Output ($Y$) defined as GDP}
    \item{All inputs and outputs normalized to initial year (0)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
    \vspace{\baselineskip}
    \pause
    \footnotesize
    \\ No known rigorous, statistical comparative studies of economic growth models for various types of economies. 
    \normalsize
  \end{center}
\end{frame}

@@

\minorSeparator{Economic Data}

\begin{frame}{Economic Data} %----------------------------
  Lucas complete
\end{frame}

@@

@@

\begin{frame}{Two Important Questions for\\Cobb-Douglas Model} %----------------------------
  \begin{center}
    \cdeEqn
  \end{center}
  \begin{itemize}
    \item{Is energy important for this economy?\\Yes, if $\gamma$ not small.}
    \vspace{\baselineskip}

@@

@@

@@

\begin{frame}{Tentative Conclusions} %----------------------------
  \begin{itemize}
  \item{Including energy in Cobb-Douglas and CES production functions is unwarranted \\
        \scriptsize
        Statistical evidence indicates that energy parameters are not significant for most countries \\
        Why? Energy is \emph{always already} embedded in capital stock ($K$) and labor ($L$), so including energy is unnecessary.
        \normalsize}
  \item{Including energy in LINEX production function is required \\
        \scriptsize
        The foundation of LINEX is an energy term.\\
        Statistical evidence indicates that LINEX parameters $a_0$ and $c_t$ are significant for most countries. \\
        LINEX fits overall are worse than Cobb-Douglas or CES.
        \normalsize}
  \vspace{\baselineskip}
  \pause

>>>>

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item{$\sim$10\%}
    \vspace{\baselineskip}
    \pause
    \item{Energy (relatively) unimportant for classical economists}
    \item{Usually ignored}
    \vspace{\baselineskip}
    \pause
    \item{And yet \dots}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Growth_Models.pdf}
  \end{center}
  \begin{itemize}
    \item{Goal: predict economic output over time}
    \item{Economic output ($Y$) is a function of ``factors of production''}
    \item{Output ($Y$) measured by GDP}
    \item{All inputs and outputs normalized to initial year: $y = Y/Y_0, e = E/E_0$, etc.}
  \end{itemize}
\end{frame}

@@

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
    \vspace{\baselineskip}
    \pause

    Extra credit: reflect dynamics of energy cost share and prices spikes.
    \vspace{\baselineskip}
    \pause

    \footnotesize
    No known rigorous, statistical comparative studies of economic growth models for various types of economies. 
    \normalsize
  \end{center}
\end{frame}

@@

\minorSeparator{Sources of Economic Data}

\begin{frame}{Economic Data: Definitions} %----------------------------
  \begin{itemize}
    \item{Output ($Y$): the real market value of the total amount of produced goods in an economy (GDP).}
    \item{Capital Stock ($K$): the real market value of the total amount of goods in an economy used in the production process of output (machines, tools, infrastructure, etc.).}
    \item{Labor: the number of workers in an economy that produce output. Labor can be treated as either the total number of workers in an economy or the total number of hours worked per year in an economy.}
  \end{itemize}
\end{frame}

\begin{frame}{Sources of Economic Data} %----------------------------
  \begin{center}
    \tiny
    \begin{tabular}{r c|c l}
      \multirow{3}{*}{WB GEP} & \multirow{3}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & $Y$, $K$ for all countries & \multirow{3}{*}{Real 2005\$} \\
                              &                          & (1980-2011: US, UK, JP; & \\
                              &                          & 1991-2011: CN, ZA, SA, IR, TZ, ZM) & \\
      \hline
      \multirow{4}{*}{WB WDI} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & Pop, Pop < 15 yo,    & \multirow{4}{*}{Total Employment} \\
                              &                          & Emp/Pop ratio > 15 yo, &                                \\
                              &                          & used to find $L$, (1991-2011: &                         \\
                              &                          & CN, ZA, SA, IR, TZ, ZM) &                         \\
      \hline
      \multirow{2}{*}{US BLS} & \multirow{2}{*}{\includegraphics[scale=0.07]{figure_other/Figure_BLS.png}} & $L$ for developed countries & Labor hours \\
                              &                          & (1980-2011: US, UK, JP)     & per year \\ 
    \end{tabular}
  \normalsize
  \end{center}
\end{frame}

@@

\begin{frame}{Comments on Single-factor Model with $Q$} %----------------------------
  \begin{itemize}
    \item{Not typically used for economic modeling. Illustrates confidence intervals.}
    \vspace{\baselineskip}

    \item{$CI_{\lambda}$ smaller than $CI_{\gamma}$ for all countries}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Two Important Questions for the\\Cobb-Douglas Model} %----------------------------
  \begin{center}
    \cdeEqn
  \end{center}
  \begin{itemize}
    \item{Is energy important for this economy?\\Yes, if $\gamma$ not small.}
    \vspace{\baselineskip}

@@

\begin{frame}{Two Important Questions for\\CES Model} %----------------------------
  \begin{center}
    \small \cesEqn \normalsize
  \end{center}
  \begin{itemize}
    \item{Is energy important for this economy?\\Yes, if $\delta$ is not large.}
    \vspace{\baselineskip}

    \item{Are we sure? (Is energy statistically significant?)\\Yes, if $CI_{\delta}$ doesn't cross 0.0.}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Two Important Questions for\\LINEX Model} %----------------------------
  \begin{center}
    \linexEqn
  \end{center}
  \begin{itemize}
    \item{Is energy important for this economy?\\Yes, by definition!}
    \vspace{\baselineskip}

    \item{Are we sure? Yes, by definition!}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Tentative Conclusions} %----------------------------
  \begin{itemize}
  \item{Including energy in Cobb-Douglas and CES production functions is unwarranted \\
        \scriptsize
        Statistical evidence indicates that energy parameters are not significant for most countries \\Why? Energy is \emph{always already} embedded in capital stock ($K$) and labor ($L$), so including energy is unnecessary.
        \normalsize}
  \item{Including energy in LINEX production function is required \\
        \scriptsize
        The foundation of LINEX is an energy term.\\Statistical evidence indicates that LINEX parameters $a_0$ and $c_t$ are significant for most countries.\\LINEX fits overall are worse than Cobb-Douglas or CES.
        \normalsize}
  \vspace{\baselineskip}
  \pause
